---
title: "Tidying and Transforming Data"
author: "Kayleah Griffen"
date: "2/20/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The objective of this assignment is to read in a csv file that has the flight delays for two airlines in a wide format. Then to use `tidyr` and `dplyr` to tidy and transform the data. Lastly, to perform an analysis to compare the arrival delays for the two airlines. 

## Tidying and Transforming Data

First, load required packages.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(kableExtra)
```

Next, read in the csv file and take a look.

```{r}
# df <- read.csv(choose.files())
df <-read.csv("https://raw.githubusercontent.com/klgriffen96/spring23_data607_hw5/main/flights.csv")
kable(head(df))  |>
  kable_styling("striped")
```

Rename headers.

```{r}
df <- df |> rename(Airline = colnames(df)[1], Status = colnames(df)[2])
```

Now, remove unnecessary blanks, fill in airlines, and ensure all are numeric.

```{r}
# Take only the rows which have values
df <- subset(df, nchar(Status) > 0)

# Fill in airline, even when it isn't there
for (i in 1:dim(df)[1]){
  temp <- nchar(df$Airline[i])
  if (temp > 0) {
    airline <- df$Airline[i]
  } else {
    df$Airline[i] <- airline
  }
}

# Make sure all values are numeric
n <- colnames(df)
for (i in 3:length(n)){
  df[[n[i]]] <- as.numeric(str_remove(df[[n[i]]], ","))
}
kable(head(df))  |>
  kable_styling("striped")
```

Perform the pivot to go from wide format to long format so that each Airline, Airport combination is a row, then back to a wide format so that the on time and delay are their own columns.

```{r}
df_long <-df |>
   pivot_longer(cols = Los.Angeles:Seattle,
               names_to = "Airport", 
               values_to = "count")

df_wide <- df_long |>
   pivot_wider(names_from = Status,
               values_from = count
   )

kable(head(df_wide))  |>
  kable_styling("striped")
```

## Analysis of Arrival Delays

To analyze the arrival delays, do a basic comparison of the overall percent of delayed flights were for each airline, then each airline by each location.

```{r}
percent_delayed <- df_wide |>
     group_by(Airline, Airport) |> 
     summarize(
         Percent_Delayed = round(`delayed`*100/(`delayed` + `on time`))
         , .groups = "drop") |>
    arrange(desc(Percent_Delayed))

overall_delayed <- df_wide |>
    group_by(Airline) |>
    summarise(
        Overall_Percent_Delayed = round(sum(`delayed`)*100/sum(`delayed` + `on time`))
    )

# Display table - head 
kable(head(percent_delayed))  |>
  kable_styling("striped")
```

```{r}
# Display overall percent 
kable(overall_delayed) |>
  kable_styling("striped")
```

Alaska airline overall delayed about 13% of it's flights and AM West overall delayed about 11% of its flights. The highest percent of flights delayed by Alaska were to San Francisco at 17% and for AM West it was also San Francisco at 29%. 

Create a bar chart showing the ratios of on-time and delayed for each location.

```{r}
ggplot(percent_delayed, aes(x=Airport, y=Percent_Delayed, fill= Airline)) + 
    geom_bar(stat="identity" , position=position_dodge())
```

The bar chart makes it clear that across all destinations, AM West had a higher percent of delayed flights than Alaska did. This is an interesting observation because overall - Alaska had a greater percentage of delays than AM West - but when you look at each destination individually you can see that AM West had a greater percentage of delays by each destination.

## Conclusion

In conclusion, I read in a csv file that had the flight delays for two airlines in a wide format. Then I used `tidyr` and `dplyr` to tidy and transform the data. Lastly, I performed an analysis to compare the arrival delays for the two airlines. 